---
title: "Slope and aspect creation with DEM"
output: pdf_document
---
# Overview

I have a DEM file which is a tiff foramt type acquired for North Western Territory in Canada for the year 2023.

The name of the data is `NWT- 2023-DEM.tif` and it contains a raster image and source note for NWT- 2023-DEM.tif


To make it easy to share this dataset I will read it, convert it to an `.RData` file, and then document it so that it can be bundled with package `CarbR`.

Load the packages needed to preprocess the data:
```{r}
#install.packages("raster")
library(raster)
```


```{r}
# Read the raster data:
Elevation <- raster("/Users/chinyereottah/Desktop/CarbR/data-raw/NWT-2023-DEM.tif")
```


```{r}
# Get a summary of the raster:
summary(Elevation)
```


The source for these data are as follow:
2023 DEM of NWT, Canada :https://search.earthdata.nasa.gov/search/?fi=ASTER


In addition, I want to fill the DEM. Filling a DEM when working with raster data ensures that the  ensure that the elevation data is continuous and there are no missing values. 
```{r}
# Check if there are any NAs (missing values) in the DEM
has_nas <- any(is.na(values(Elevation)))

if (has_nas) {
  # Fill the gaps using a simple interpolation method (linear interpolation)
  filled_dem <- raster::interpolate(Elevation, method = "bilinear")
} else {
  filled_dem <- Elevation
}
```

Save to data folder:
```{r}
usethis::use_data(Elevation, overwrite = TRUE)

```

```{r}
#writeRaster(elevation_raster, filename = "NWT-2023-DEMfill.tif", format = "GTiff")
```
