---
title: "CarbR-vignette"
output: rmarkdown::html_vignette
abstract: "Ensuring the reproducibility of research is a crucial aspect of scholarly endeavors. One key element in achieving reproducibility is the comprehensive documentation of the processes employed, encompassing both data processing and subsequent analyses. This vignette is tailored for the CarbR package, a valuable tool designed for generating slope and aspect maps using digital elevation models (DEM). The package also guides users through the process of DEM filling and presents a framework for conducting analyses in a reproducible manner.In addition to elucidating the steps involved in creating slope and aspect maps, this vignette underscores the significance of transparently documenting the entire analytical workflow. To enhance accessibility, the DEM utilized in the examples presented herein will be seamlessly integrated into the CarbR package."
keywords: "R packages; slope, aspect, DEM"
author: "Chinyere Ottah"
date: "07 November 2023"
vignette: >
  %\VignetteIndexEntry{CarbR-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Introduction
CarbR is an R package for deriving slope and aspect maps using the `calculate_slope_aspect` function. This function takes a raster object representing the Digital Elevation Model (DEM) as input and returns two rasters representing slope and aspect. The package is designed to assist in various analyses, providing a cost-free solution using R. It also facilitates easy distribution and reproducibility among researchers, teammates, and decision-makers.
## Overview

CarbR is suitable for researchers, students, and decision-makers interested in surface analysis to understand topographic influences. This can be applied to various scenarios such as wildfire behavior, flooding, and other disturbances where slope and aspect play crucial roles.

## Sample Data

The sample data included in this package was collected from the North Western Territory (NWT) in Canada for the year 2023. You can download the sample data from [GitHub](https://github.com/ChinyereRuth/CarbR) or directly from the [USGS webpage](https://search.earthdata.nasa.gov/search/granules?projectId=2435656526).


## CarbR Function: Aspect and Slope

The `calculate_slope_aspect` function in CarbR is used to create aspect and slope maps. Given a raster DEM as input, the function returns rasters for slope and aspect in degrees. Slope indicates the steepness and inclination of a surface, while aspect describes the direction of the slope (North, East, West, South).

Not only does the function enable the fast generation of aspect and slope maps, but it also facilitates the derivation of new maps for various analyses.

The CarbR package provides two terrain surface outputs:

- **Slope:** Indicates the steepness and inclination of a surface area.
- **Aspect:** Describes the direction of the slope and is measured in degrees (North, East, West, and South).

## Example Usage

Here are concrete examples of using the `calculate_slope_aspect` function in CarbR:

### Preliminaries

Ensure that R and RStudio are installed. Install the CarbR package from GitHub:


# install.packages("remotes")
# remotes::install_github("ChinyereRuth/CarbR")

Also, install and load the required raster package:

```
library(raster)
```
## Workflow for this package

Firstly, we have to load the elevation data 

To load the data, use the function `data()`:
```
data("Elevation")
```

To inspect the object, use the function `summary()`. 

```
summary(Elevation)
```
This will provide a summary estimate based on a sample of 1e+05 cells (0.77% of all cells).

## Quality checks on data 
Next, we check for missing values for the raster data. if there are missing values, We will fill the DEM. Filling a DEM when working with raster data ensures that the  ensure that the elevation data is continuous and there are no missing values. 

```
# Check if there are any NAs (missing values) in the DEM
has_nas <- any(is.na(values(Elevation)))

if (has_nas) {
  # Fill the gaps using a simple interpolation method (linear interpolation)
  filled_dem <- raster::interpolate(Elevation, method = "bilinear")
} else {
  filled_dem <- Elevation
}
```
Next, we can save the filled DEM and use it to calculate slope and aspect using the slope aspect function. 
# usethis::use_data(Elevation, overwrite = TRUE)

To save the DEM, use the following command line: 
```{r}
# usethis::use_data(Elevation, overwrite = TRUE)
```

Next, we use the slope and aspect function from the CarbR package to calculate the aspect and slope using the saved DEM as input

```{r}
library(raster)
data("Elevation")
#summary(Elevation)
library(CarbR)
CarbR::calculate_slope_aspect(Elevation)
```
Next, we get the result of the function and call the result.
```{r}
result <- CarbR::calculate_slope_aspect(Elevation)
slope_raster <- result$slope
aspect_raster <- result$aspect
```

We can plot the result to see the aspect and slope 
```{r}
plot(slope_raster, main="Slope")
plot(aspect_raster, main="Aspect")
```

## Conlusion
From the result, we see the slope and aspect shown in degrees. This result is very important for carbon loss modelling, especially from fire. Steep slopes from this result  increase the spread of fire as fires tend to progress uphill because the heat and flames have already heated up the vegetation as the fire moves uphill, making it easier for them to burn and spread. For the aspect which is important for  determining the direction of the slope. South-facing slopes tend to have more sunlight, especially within the northern hemisphere, making the fuels drier and more susceptible to burning. Therefore, the CarbR package is very useful for assessing fire disturbances, including emissions and other standing and non-standing disturbances. 
